---
title: "Generalized Additive Models for Actuaries"
author: "Nathan Cornwell"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
site: bookdown::gitbook
documentclass: book
bibliography: book.bib
biblio-style: apalike
link-citations: yes
github-repo: null
description: An Introduction to Generalized Additive Models for Health Actuarial Business
  Problems.
---

# Introduction {.unnumbered}

[LinkedIn](https://www.linkedin.com/in/nathan-cornwell-asa-maaa-01170011b/)

[Email](nathan.cornwell@optum.com)

## Our Workspace

[R/RStudio](https://posit.co/download/rstudio-desktop/)

[Github Repo](https://github.com/Theaetetos/gams-for-actuaries)

[Github Page](https://theaetetos.github.io/gams-for-actuaries/)

## Preliminaries

### Packages

Here are all of the packages we will need for this presentation.
You only need to run the first code chunk once for each installation of R you have, while you need to run the second every time you start a new R session.

```{r, eval=FALSE}
install.packages(
  c('data.table', 'ggplot2', 'gratia', 'gridExtra', 'mgcv',
    'parallel', 'scales')
)
```

```{r, message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(gratia)
library(gridExtra)
library(mgcv)
library(parallel)
library(scales)
```

### One of Our Datasets

We will use the diamonds dataset in the first lesson. 
The rest of the data is modified, masked, and randomized real data from projects I've done.

```{r}
data('diamonds')
```

### Branding Code

Finally, this is some branding stuff that I use to make my plots look nice and comply with Optum and UHC branding guidelines.
You don't need it outside of the code in the presentation, but you can use it as a template for customizing your own plots, if you wish.

```{r}
primary_colors <- list(`Optum Orange` = '#FF612B')
sec_colors <- list(White = '#FFFFFF', `Sky Blue` = '#D9F6FA',
                   `Warm White` = '#FBF9F4')
type_colors <- list(`Dark Blue` = '#002677', `Dark Gray` = '#5A5A5A')
viz_colors <- list(Violet = '#422C88', Iris = '#8061BC',
                   Lagoon = '#007C89', Rainwater = '#6FC1B1',
                   Strawberry = '#A32A2E', Apple = '#D13F44')
state_colors <- list(`Green Success` = '#007000',
                     `Gold Callout` = '#F5B700',
                     `Red Alert` = '#C40000')
uhc_colors <- list(Blue = '#002677', White = '#FFFFFF',
                   `Bright Blue` = '#00BED5', Gold = '#F5B700',
                   Orange = '#FF681F', `Bright Blue 20%` = 'CCF2F7',
                   `Bright Blue 40%` = '#99E5EE',
                   `Dark Gray` = '#5A5A5A', `Light Gray` = '#F4F4F4',
                   `Medium Gray` = '#DADBDC', Hyperlink = '#196ECF')
colors <- list(Primary = primary_colors, Secondary = sec_colors,
               Typography = type_colors, Visualization = viz_colors,
               State = state_colors, UHC = uhc_colors)

scale_discrete_optum <- function(aesthetics, ...){
  discrete_scale(
    aesthetics = aesthetics,
    scale_name = 'optum',
    palette = grDevices::colorRampPalette(
      as.character(colors$Visualization[c(4, 3, 1, 2, 6, 5)])
    ),
    ...
  )
}

theme_titles <- theme(
  plot.title = element_text(hjust = .5),
  plot.subtitle = element_text(hjust = .5)
)
theme_rotx <- theme(axis.text.x = element_text(angle = 90))
theme_optum_plot <- theme(
  text = element_text(color = colors$Typography$`Dark Gray`),
  panel.background = element_rect(
    fill = colors$Secondary$White,
    color = colors$Typography$`Dark Gray`
  )
)

mytheme <- theme_optum_plot + theme_titles
```
