---
title: "Generalized Additive Models for Actuaries"
author: "Nathan Cornwell"
date: "`r Sys.Date()`"
site: bookdown::gitbook
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
github-repo: 
description: "An Introduction to Generalized Additive Models for Actuarial Business Problems."
---

# Introduction {#intro}

This is just some branding stuff that I use to make my plots look nice and comply with Optum and UHC branding guidelines.
You don't need it outside of the code in the presentation, but you can use it as a template for customizing your own plots, if you wish.

```{r}
primary_colors <- list(`Optum Orange` = '#FF612B')
sec_colors <- list(White = '#FFFFFF', `Sky Blue` = '#D9F6FA',
                   `Warm White` = '#FBF9F4')
type_colors <- list(`Dark Blue` = '#002677', `Dark Gray` = '#5A5A5A')
viz_colors <- list(Violet = '#422C88', Iris = '#8061BC',
                   Lagoon = '#007C89', Rainwater = '#6FC1B1',
                   Strawberry = '#A32A2E', Apple = '#D13F44')
colors <- list(Primary = primary_colors, Secondary = sec_colors,
               Typography = type_colors, Visualization = viz_colors)

# plotting
scale_discrete_optum <- function(aesthetics, ...){
  discrete_scale(
    aesthetics = aesthetics,
    scale_name = 'optum',
    palette = grDevices::colorRampPalette(
      as.character(colors$Visualization[c(4, 3, 1, 2, 6, 5)])
    ),
    ...
  )
}

theme_titles <- theme(
  plot.title = element_text(hjust = .5),
  plot.subtitle = element_text(hjust = .5)
)
theme_rotx <- theme(axis.text.x = element_text(angle = 90))
theme_optum_plot <- theme(
  text = element_text(color = colors$Typography$`Dark Gray`),
  panel.background = element_rect(
    fill = colors$Secondary$White,
    color = colors$Typography$`Dark Gray`
  )
)

mytheme <- theme_optum_plot + theme_titles
```

Here are all of the packages we will need for this presentation.
You only need to run the first code chunk once for each installation of R you have, while you need to run the second.

```{r, eval=FALSE}
install.packages(
  c('data.table', 'ggplot2', 'gratia', 'gridExtra', 'insuranceData',
    'mgcv', 'parallel', 'scales')
)
```

```{r}
library(data.table)
library(ggplot2)
library(gratia)
library(gridExtra)
library(insuranceData)
library(mgcv)
library(parallel)
library(scales)
```

Finally, these are two of the datasets we will use in the examples.

```{r}
data('diamonds')
data('dataCar')
```
